---
title: "Homework 3"
subtitle: "Research Methods, Spring 2025"
author: "Genevieve DeBell"
format:
  pdf:
    output-file: "debell-g-hwk3-2"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| include: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes, fixest,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, kableExtra, modelsummary,
               kableExtra, tinytex)
```
```{r}
#| include: false
#| eval: true 
load("Hwk3_workspace.RData")
load("Hwk3datacode_workspace.RData")

```
My answers to the homework questions are described below. The GitHub repository for this work is available [here](https://github.com/genevievedebell/hwk3). Enjoy!

# 1. Present a bar graph showing the proportion of states with a change in their cigarette tax in each year from 1970 to 1985.

```{r}
#| echo: false
## Identify years with tax changes 
tax_changes <- final.data %>%
  filter(Year >= 1970, Year <= 1985) %>%
  arrange(state, Year) %>%
  group_by(state) %>%
  mutate(tax_change = tax_state != lag(tax_state, default = first(tax_state))) %>%
  ungroup()

## calculate the proportion of states that had a tax change in each year
change_proportions <- tax_changes %>%
  group_by(Year) %>%
  summarize(proportion = mean(tax_change, na.rm = TRUE))

## plot the results
library(ggplot2)

ggplot(change_proportions, aes(x = Year, y = proportion)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Proportion of States with Cigarette Tax Changes (1970-1985)",
       x = "Year",
       y = "Proportion of States") +
  theme_minimal()
```

# 2. Plot on a single graph the average tax (in 2012 dollars) on cigarettes and the average price of a pack of cigarettes from 1970 to 2018.

```{r}
#| echo: false
## Get the CPI value for 2012
cpi_2012 <- cpi.data %>% filter(Year == 2012) %>% pull(index)

## Convert tax and price to 2012 dollars
final.data <- final.data %>%
  mutate(price_real = cost_per_pack * (cpi_2012/index),
         tax_real = tax_dollar * (cpi_2012/index))

## Filter for years 1970 to 2018
plot.data <- final.data %>%
  filter(Year >= 1970 & Year <= 2018) %>%
  group_by(Year) %>%
  summarize(avg_tax = mean(tax_real, na.rm = TRUE),
            avg_price = mean(price_real, na.rm = TRUE))

## Plot the data
library(ggplot2)

ggplot(plot.data, aes(x = Year)) +
  geom_line(aes(y = avg_tax, color = "Average Tax (2012$)"), size = 1.2) +
  geom_line(aes(y = avg_price, color = "Average Price (2012$)"), size = 1.2) +
  labs(title = "Average Tax and Price of Cigarettes (1970-2018, Adjusted to 2012$)",
       x = "Year",
       y = "Dollars (2012 Adjusted)",
       color = "Legend") +
  theme_minimal()
```

# 3. Identify the 5 states with the highest increases in cigarette prices (in dollars) over the time period. Plot the average number of packs sold per capita for those states from 1970 to 2018.

```{r}
#| echo: false
## Ensure 'Year' is an integer
final.data <- final.data %>%
  mutate(Year = as.integer(Year))

## Calculate price difference (2012 dollars)
price_diff <- final.data %>%
  filter(Year %in% c(1970, 2018)) %>%
  select(state, Year, price_real) %>%
  pivot_wider(names_from = Year, values_from = price_real, names_prefix = "Year_") %>%
  mutate(price_increase = Year_2018 - Year_1970) %>%
  arrange(desc(price_increase))

## Identify the top 5 states with the highest price increase
top_5_states <- price_diff %>%
  slice_max(order_by = price_increase, n = 5) %>%
  pull(state)

## Filter data for these 5 states and sales per capita
sales_data <- final.data %>%
  filter(state %in% top_5_states, Year >= 1970 & Year <= 2018) %>%
  select(state, Year, sales_per_capita)

## Plot the sales per capita for these states
ggplot(sales_data, aes(x = Year, y = sales_per_capita, color = state)) +
  geom_line(size = 1.2) +
  labs(title = "Average Packs Sold Per Capita (1970-2018)",
       subtitle = "For the 5 States with the Highest Increase in Cigarette Prices",
       x = "Year",
       y = "Packs Sold Per Capita",
       color = "State") +
  theme_minimal()
```

# 4. Identify the 5 states with the lowest increases in cigarette prices over the time period. Plot the average number of packs sold per capita for those states from 1970 to 2018.

```{r}
#| echo: false
## Ensure 'Year' is an integer
final.data <- final.data %>%
  mutate(Year = as.integer(Year))

## Calculate price difference (2012 dollars)
price_diff <- final.data %>%
  filter(Year %in% c(1970, 2018)) %>%
  select(state, Year, price_real) %>%
  pivot_wider(names_from = Year, values_from = price_real, names_prefix = "Year_") %>%
  mutate(price_increase = Year_2018 - Year_1970) %>%
  arrange(price_increase)  # Sort in ascending order

## Identify the 5 states with the lowest price increase
bottom_5_states <- price_diff %>%
  slice_min(order_by = price_increase, n = 5) %>%
  pull(state)

## Filter data for these 5 states and sales per capita
sales_data <- final.data %>%
  filter(state %in% bottom_5_states, Year >= 1970 & Year <= 2018) %>%
  select(state, Year, sales_per_capita)

## Plot the sales per capita for these states
ggplot(sales_data, aes(x = Year, y = sales_per_capita, color = state)) +
  geom_line(size = 1.2) +
  labs(title = "Average Packs Sold Per Capita (1970-2018)",
       subtitle = "For the 5 States with the Lowest Increase in Cigarette Prices",
       x = "Year",
       y = "Packs Sold Per Capita",
       color = "State") +
  theme_minimal()
```

